\lab{MongoDB}{MongoDB}
\objective{A very basic, non-comprehensive introduction to MongoDB.}
\label{lab:mongodb}

\section*{NoSQL Databases}
Relational databases, such as SQL, were the most popular databases of the last decade.  These databases rely on the data having relational attributes, meaning that each item in the database has the same attributes.  As time passed and needs changed, relational databases became impractical for some sets of data.  Sometimes the relation model is too structured.  Each item may not have the same attributes.  For example, a salamander and an apple both have attributes of size and color, but an apple does not need a gender attribute and a salamander does not need a ripeness attribute.  Relational databases store items with the same attributes, but if we want to store a salamander and an apple in the same database, we need a different type of database, hence the need for non-relational databases.

A new family of databases arose that attempted to solve the problem.
Instead of designing a new relational database to meet every need, non-relational databases were created that can adapt for specific scenarios.

MongoDB is such a database.  Several other databases, such as Cassandra, Redis, and Neo4j serve similar purposes.  In this lab, we will focus on MongoDB.

\section*{MongoDB}
MongoDB is a document database.
It is best suited for storing data that does not have a fixed schema.  Each MongoDB database is made up of collections of one or more documents.
These documents are a special type of JSON object called BSON (Binary JSON).  MongoDB has both a commandline interface and Python bindings.
This lab will use the official supported Python bindings, pymongo.


\begin{info}
Make sure you have a MongoDB server running before trying to connect to a database via Python.
It is best to be running the latest stable version of MongoDB (at least version 2.6).
\end{info}

You can run a MongoDB server as follows:
\begin{lstlisting}[language=]
$ mongod --dbpath ./data/db
\end{lstlisting}

Once you have the server running, you can use it in a Python envirnoment as follows:

\begin{lstlisting}
from pymongo import MongoClient

# Create an instance of a client
# Connect on the default host and port
mc = MongoClient()
\end{lstlisting}

\subsection*{Basic Example}
Suppose you're running a general store. You have all sorts of inventory, such as food, clothing, tools, toys, etc.  There are some attributes that every item has, like name, price, and producer.  Then there are attributes held by only some items, for example color, weight, gluten-freedom.  A SQL database would have to be full of mostly-blank rows, which is inefficient.  More importantly, as you add new inventory, you'll run across new attributes. With SQL, you'd have to restructure and rebuild the whole database each time this happened. For MongoDB, this isn't a problem. That's because it doesn't use relation tables. Each item is a JSON-like object (similar to a Python dictionary), and thus can contain whatever attributes are relevant to the specific item.

\section*{Creating and Removing Collections and Documents}
A database stores collections, and a collection stores documents. This is the basic heirarchy of MongoDB.  Each database can have several collections, each with their own documents.  We need to create a database that will hold our collections.  Imagine we have a set of paper documents.  We put the documents into folders (collections), and the folders into a filing cabinet (the database).  When we need to add another collection, we simply create a reference to it.  The new collection will not actually be created until we add documents to it, just as we would not put a folder into the filing cabinet until we have a document to be put in the folder.
\begin{lstlisting}
# Create a new database
db = mc.db1

# Create a new collection
col = db.collection1
\end{lstlisting}

Documents in MongoDB are represented as JSON-like objects.
MongoDB databases don't adhere to a set schema.
Each document can have its own fields.
\begin{lstlisting}
col.insert({'name': 'Jack', 'age': 23})
col.insert({'name': 'Jack', 'age': 22, 'student': True, 'classes': ['Math', 'Geography', 'English']})
x = col.insert({'name': 'Jill', 'age': 24, 'student': False})
\end{lstlisting}
We can check to see if the insert was successful by calling \li{x.is_valid(x)}.

Now we can update documents in a collection using the update.
Note that a simple update acts like a replace.
\begin{lstlisting}
col.update({'name': 'Jack'})
\end{lstlisting}



\section*{Querying for Documents}
MongoDB uses a \emph{query by example} paradigm for querying.
This means that when you query, you provide an example that the database uses to match with other documents.
\begin{lstlisting}
# Querying methods return a Cursor object which iterates through result set
r = col.find({'name': 'Jack'})
\end{lstlisting}
This query will return all documents in the collection that have the value `Jack' in the `name' field.


\subsection*{Query Operators}
There are several special operators that we can use to define conditions in a query.
\begin{table}
\begin{tabular}{|l|l|}
\hline
Operator & Description \\
\hline
\$lt, \$gt & $<$, $>$ \\
\$lte & $\leq$, $\geq$ \\
\$in, \$nin & Match any value in, not in an array, respectively \\
\hline
\$or & Logical OR \\
\$and & Logical AND \\
\$not & Logical negation \\
\$nor & Logical NOR (condition fails for all clauses) \\
\hline
\$exists & Match documents with specific field \\
\$type & Match documents with values of a specific type \\
\hline
\$all & Match arrays that contain all queried elements \\
\end{tabular}
\caption{MongoDB query operators}
\label{table:queryoperators}
\end{table}
These query operators are used as keys and the queries are values.
\begin{lstlisting}
f = list(col.find({'age': {'$lt': 24}, 'classes': {'$in': ['Art', 'English']}}))
\end{lstlisting}

Understand that MongoDB is not a relational database, therefore there is no concept of a join.
This also means that we cannot define database relationships between documents.
We can associate two documents by including a field that contains the unique ObjectID of the other document.
When we request one document, we see it has an ObjectID, and then we run a second query to get the other document.
Any ``relational'' things must be handled by the developer.
This means that a document needs to contain all the information needed to find or retrieve it again.